\Large\textbf{Advection versus diffusion in brain ventricular transport}

### 2025-04-04 paper/main.tex
revision edition

\begin{abstract}
Cerebrospinal fluid (CSF) is integral to brain function, as the fluid
provides mechanical support for the brain and the flow of CSF helps
distribute nutrients, neurotransmitters and metabolites throughout the
central nervous system. CSF flow is partly driven by beating of motile
cilia, hair-like structures that line the ependymal cell layer of the
walls of the brain ventricles. Despite the physiological importance of
CSF flow, the underlying mechanisms of cilia-mediated flow and
transport in the brain ventricles remain yet to be comprehensively
resolved. This study analyzes and evaluates the role of motile cilia
in brain intraventricular CSF flow and transport. The geometry of the
zebrafish larval brain ventricles, for which detailed knowledge of
cilia properties and CSF motion velocities and frequencies exists, was
used to develop finite element methods for modelling flow and
transport. The computational model is validated by in-vivo experiments
that monitor transport of a photoconvertible protein secreted in the
brain ventricles. Our results show that while cilia contribute to
advection of large particles, diffusion plays a significant role in
the transport of small solutes. We also demonstrate how cilia location
and the geometry of the ventricular system impact solute
distribution. Altogether, this work presents a computational framework
that can be applied to other ventricular systems, together with new
concepts of how molecules are transported within the brain and its
ventricles.

@@

\noindent\small\hspace{.8725cm}\textbf{Keywords} cerebrospinal fluid, brain ventricles, cilia, transport, finite elements

@@

% I4: How do we address the problem and what do we find out?
In this study, we combined computational and experimental approaches to investigate
the role of motile cilia in intraventricular CSF flow and transport.
We model the flow and transport by partial differential equations and solve
them numerically with finite element methods to compute flow patterns and
solute concentration dynamics. A geometry representing the zebrafish larval brain ventricles
was used, for which we have detailed knowledge of cilia properties and CSF motion velocities
and frequencies~\cite{Olstad2019CiliaryDevelopment}. 
Notably, this early developmental stage allows us to simplify our model
by omitting CSF secretion, which starts at later stages~\cite{Jeong2024TheZebrafish},
while still being well-conserved with mammals~\cite{DGama2025MotileBrain,
Olstad2019CiliaryDevelopment, Ringers2020Role, DGama2021Diversity, jurisch2020radial}. 
The computational model is validated against experimental data by monitoring in vivo
the dynamics of a photoconvertible fluorescent protein Dendra2
expressed within the brain ventricles.
Our results show that diffusion plays a major role in the transport of small solutes,
whereas cilia are paramount for the movement of larger particles.
We also found that the location of cilia, as well as changes to the ventricular geometry,
largely impact solute distribution.
In summary, our work presents a computational
framework that can be applied to other
ventricular systems in the future, together with new concepts
of how cilia distribute molecules and
particles within the brain ventricles.

@@

Assuming that viscous forces are dominating the intraventricular CSF flow,
we model the flow by the Stokes equations.
The net forces of motile cilia acting on the CSF are modeled as a steady,
bidirectional tangential traction acting on parts of the ventricular wall
(\Crefpanels{fig:fig1}{d}), and we assume free velocity slip at the walls.
Simulations show that this traction alone
leads to a partial compartmentalization of the ventricular system,
with large-scale vortex structures in the anterior, middle and
posterior ventricles (\Crefpanels{fig:fig1}{e}). The highest flow
speeds ($26.7$ \textmu m/s) occur in the vicinity of the dorsal cilia
in the middle ventricle. To model the cardiac pulsations, we impose a
pulsatile pressure difference between the anterior and posterior
boundaries (\Crefpanels{fig:fig1}{d, f}). The resulting flow
is pulsatile and directional (\Crefpanels{fig:fig1}{g}), and
in the absence of the cilia, the vortex structures vanish. The
velocity magnitude peaks at $4.8$ \textmu m/s. CSF initially flows
rostrocaudally before changing direction to caudorostral in the middle
of the cardiac cycle.

@@

Ciliary motion is integral to physiological CSF flow, and absence of
cilia has been associated with defective brain development and various
pathologies~\cite{Eichele2020Cilia-drivenVentricle,
  Olstad2019CiliaryDevelopment, Faubel2016Cilia-basedVentricles,
  Guirao2010CouplingCilia, Afzelius2004CiliaRelatedDiseases,
  Sawamoto2006NewBrain, Yoshiba2014RolesSymmetry,
  Hirokawa2006NodalAsymmetry}.  As observed both experimentally and
computationally, absence of motile cilia alters CSF flow patterns.  We
next investigate the implications of these changes on intraventricular
transport. We experimentally measured the movement of proteins by
monitoring photoconverted Dendra2 signal in control zebrafish and
motile-cilia mutant \emph{schmalhans} (\emph{smh}) zebrafish.  In ROIs
2 and 4, which quantify Dendra2 distribution downstream of the
photoconversion site, we observed a slower increase of fluorescence
intensity (\Crefpanels{fig:fig4}{a}): the mean times-to-threshold in
the \emph{smh} mutants were $45\%$ (ROI 2) and $25\%$ (ROI 4) higher
compared to controls (\Crefpanels{fig:fig4}{b}). We found no
significant differences between \emph{smh} mutants and controls in the
Dendra2 distribution in ROIs 3, 5, and 6.

In the computational model, we represent the \emph{smh} mutants by
considering transport of Dendra2 with CSF flow driven by cardiac
pressure pulsations only, representing the absence of motile
cilia. Comparing the in silico mutants with transport predictions in
the presence of both cardiac pulsations and motile cilia, we observe
noticeably delayed dynamics in ROIs 2, 4 and 5, but little change for
ROIs 3 and 6 (\Crefpanels{fig:fig4}{c}). Compared to experimental
data, the control and mutant scenarios differed more for the
simulations, with the times-to-threshold in ROI 2 and 4 being 81\% and
28\% higher in the absence of cilia-mediated flow. Moreover, the
predicted mean concentrations in ROI 5 deviate between the two
scenarios, which is not the case in the experiments. Altogether, our
findings identify that, for smaller proteins such as Dendra2,
diffusion plays an important role for solute movement to regions
upstream in the cilia-mediated flow fields or to distant ventricles,
while the cilia affect the downstream mean concentrations.
\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{graphics/figure4_compare_control_mutant.png}
    \caption{\textbf{a.} Relative change in fluorescence intensity $\Delta F(t)$
    in each region of interest (ROI) for control (black lines, $n=8$)
    and \emph{smh} mutant (red lines, $n=6$) fish. 
    Bold lines are mean values of the cohorts and shaded regions cover one standard devation errors
    in the means.
    \textbf{b.} Times when $\Delta F(t)$ first exceeds a threshold value,
    equal to 0.25 (ROIs 2--4) or 0.10 (ROIs 5, 6), for control (gray dots) and
    \emph{smh} mutant (red dots) fish.
    The thick black lines denote mean values of the datasets. The $p$-values
    were calculated using a Mann-Whitney U test with a confidence level of 95\%, with 
    the null hypothesis that the two distributions of control and mutant data are 
    from the same population.
    \textbf{c.} Mean concentration $\cbar(t)$ in each ROI,
    simulated using the baseline model (black lines) and the cardiac-induced/no-cilia model (red lines).
    The horizontal gray lines mark the threshold value $\hat{c}$, and the time
    $\hat{t}$ when the threshold is first exceeded is reported in the plot legends.
    Note that in both panels \textbf{a} and \textbf{c}, the vertical axes have
    different limits in each plot.}
    \label{fig:fig4}
\end{figure}

@@

We previously characterized three distinct ciliated cell lineages in
the zebrafish larval brain~\cite{Olstad2019CiliaryDevelopment,
  DGama2025MotileBrain}.  To identify the effects of these lineages on
CSF flow and transport, we investigate region-specific cilia paralysis
with the computational model.  Three cilia-covered regions were
paralyzed separately: the anterior ventricle, and the dorsal and
ventral parts of the middle ventricle (\Crefpanels{fig:fig5}{a}).  To
simulate regional cilia paralysis, we remove the tangential traction exerted by
the cilia in the respective region.
%
%% the cilia tangential traction boundary condition is for each case
%% changed to a free-slip boundary condition in the paralyzed region,
%% prescribing zero tangential forces to represent absence of cilia
%% motion.
%

Removing cilia in the dorsal part of the middle ventricle has a strong
effect on the CSF velocities: the maximum velocity magnitude drops
from 27.9 \textmu m/s to 7.9 \textmu m/s. In vivo, these dorsal cilia
generate the strongest flow at this developmental stage of the
fish~(\Crefpanels{fig:fig1}{c})~\cite{Olstad2019CiliaryDevelopment}.
In contrast, only removing cilia in the anterior ventricle or the
ventral region of the middle ventricle has minimal impact, and results
in maximum speeds of 27.9 \textmu m/s and 27.4 \textmu m/s,
respectively.  We remark that out of the total tangential forces,
ventral cilia in the middle ventricle contribute by 34\%, while the
dorsal cilia exert 52\% of the total tangential
force~(\Crefpanels{fig:fig5}{b}). \mer{Thus, the CSF velocities are
  affected not only by the magnitude of the applied forces, but result
  from a nontrivial interplay between force, directionality and
  location.}
%% The magnitude of these force
%% contributions are thus more similar than the difference in their
%% impact on the maximum velocity magnitudes.

Turning to transport of Dendra2, we again note that removal of the
dorsal cilia has the most pronounced effect. All mean concentration
profiles (except ROI 1, where we impose the photoconversion curve)
change noticeably (\Crefpanels{fig:fig5}{c}). Compared to the baseline
model results, we observe less solute distribution in the middle and
anterior ventricles (ROIs 2--5), and more rapid transport to the
posterior ventricle (ROI 6). On the other hand, removal of anterior
cilia does not alter the mean concentrations
(\Crefpanels{fig:fig5}{c}). Finally, without ventral cilia in the
middle ventricle, more solute spreads towards the anterior ventricle
(ROI 5) and less solute spreads towards the posterior ventricle (ROI
6).
\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{graphics/figure5_compare_cilia_modifications.png}
    \caption{\textbf{a.} The three separate cilia regions for which cilia paralysis is simulated
    by imposing a zero tangential traction boundary condition in the respective region.
    \textbf{b.} The total tangential force applied to the boundary with the tangential traction
    boundary condition in each of the three separate ciliated regions, as well as the total force
    (the sum of the three regions), calculated with \Cref{eq:tangential_force_calculation}. 
    \textbf{c.} Mean concentration $\cbar(t)$ for the six regions of interest (ROIs) in the three 
    paralysis scenarios considered. The region named in the legend is the region
    where we modify the cilia tangential traction boundary condition to a free-slip condition
    in the CSF flow model.}
    \label{fig:fig5}
\end{figure}

\subsection{Modification of ventricular geometry impacts solute distribution}
Variations in ventricular morphology impact CSF flow
patterns~\cite{Olstad2019CiliaryDevelopment}.  To assess how these
variations influence solute distribution, we consider four alternative
ventricular geometries: (i) constriction of the anterior-middle
ventricular duct (66 \% reduction in cross-sectional area);
(ii) constriction of the middle-posterior duct (33 \% reduction);
(iii) middle ventricle shrunk in the lateral direction (43 \% reduction);
and (iv) all three combined (\Crefpanels{fig:fig6}{a, b}).
%
%% one where the duct connecting the anterior and middle ventricles is constricted
%% (66 \% reduction in cross-sectional area);
%% one where the duct connecting the middle and posterior ventricles is constricted
%% (33 \% reduction);
%% one where the middle ventricle is shrunk in the lateral direction (43 \% reduction);
%% and the combination of these three modifications (\Crefpanels{fig:fig6}{a, b}).
%
%
In these four modified geometries, the total volume is reduced by
3.9\%, 4.1\%, 1.4\% and 9.4\%, respectively, compared to the original
geometry. \mer{MER: HH adds a sentence on changes in tangential
  traction(\Crefpanels{fig:fig6}{d}). This is notwithstanding a
9.5\% reduction in ventral cilia force magnitude
(\Crefpanels{fig:fig6}{d}), which are the cilia that propel CSF and solute towards ROI 6.)}

Constricting the connection between the anterior and middle ventricle
reduces solute transport to the anterior ventricle (ROI 5), and the
mean concentration in ROIs 2--4 and 6 slightly increases, compared to
the original geometry (\Crefpanels{fig:fig6}{c}). \mer{We thus note
  that restricting the ...} Constricting the middle-posterior duct
mainly delays the transport into the posterior ventricle (ROI 6),
while \mer{gently} accelerating transport towards the anterior
ventricle (ROI 5). Faster transport to the anterior ventricle (ROI 5)
and slower transport to the lower middle and posterior ventricle
(ROI3, ROI 6) is also the case for the geometry with a shrunk middle
ventricle. \mer{Interestingly, this reduction in the transport to the
  posterior ventricle (ROI 6) happens in spite of a 6.6\% increase in
  the ventral cilia forces.}
%
%For the mesh where all three ventricles are modified,
When combining all three modifications,
we observe higher retention of solute in the middle ventricle (ROIs 2--4),
while the mean concentration is reduced in both the anterior
and the posterior ventricles (ROIs 5 and 6)s.

To summarize, these results show that the ventricular geometry and the
cross-sectional area of the ventricular ducts clearly impact solute
transport, with notable differences in transport resulting from modest
geometry modifications. These findings may explain the larger
differences in the cilia-deficient computational model results
compared to the experimental data in \Cref{fig:fig4}, \mer{as the
computational geometry is not an exact representation for the
individual zebrafish ventricles.}

\lyng{We previously identified large interindividual variation in ventricular size and
CSF velocities between larva at given developmental stages and throughout
development~\cite{Olstad2019CiliaryDevelopment, DGama2025MotileBrain, Jeong2024TheZebrafish}.
Our results suggest that these two parameters, CSF velocities and ventricular size,
affect each other.}
\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{graphics/figure6_compare_modified_geometries.png}
    \caption{\textbf{a.} The original computational mesh of the ventricular geometry.
    \textbf{b.} The four modified versions of the computational geometry. Compared
    to the original mesh, the volumes of the deformed meshes are reduced by 3.9\%, 4.1\%, 1.4\% 
    and 9.4\%.
    \textbf{c.} Mean concentration $\cbar(t)$ simulated with the original and modified geometries
    in the 6 ROIs.
    \textbf{d.} A comparison of the tangential cilia force in the three cilia regions
    and the total force applied, calculated with \Cref{eq:tangential_force_calculation}.
    In each region, the force $F_{\mathrm{original}}$ 
    applied on the original mesh is the baseline of 100\%,
    to which the forces $F_i$ of the other mesh versions are compared, where $i$ denotes
    the cases fore-shrunk, middle shrunk, mid-hind shrunk, and the combination of the three.
    The bars for the four shrunk scenarios display $F_i/F_{\mathrm{original}}\times 100\%$.}
    \label{fig:fig6}
\end{figure}

@@

Using the simulated CSF flow, we modeled solute transport by an advection-diffusion equation.
The transport results compared well with data from in vivo experiments in zebrafish.
Then, by using a computational model as well as zebrafish mutants with paralyzed cilia,
we identified that even though advection by cilia is dominant,
diffusion plays a significant role in protein transport.
Finally, we investigated how local cilia paralysis and ventricular geometries
affect solute distribution. Altogether, we present not only new findings on the role of
diffusion and advection in intraventricular transport, but also provide a
computational framework to study CSF flow and transport in ventricular geometries.

\subsection{The roles of advection and diffusion in intraventricular solute transport}
To investigate the roles of advection and diffusion in solute transport, we quantified transport
based on changes in fluorescence intensity $\Delta F$ (experiments) and
mean concentration $\cbar$ (simulations) over time. Our results convincingly show that
intraventricular transport of solute depends on a balance of advection and diffusion.
While advection is the strongest transport mechanism on a global scale for
Dendra2 proteins (global Péclet number 13), diffusion is balancing out
advection for shorter distances and for regions without motile cilia activity.
For instance, for the larger extracellular vesicles, no solute reached the (distant)
posterior ventricle (a distance of $L = 200$ \textmu m) within the simulation time (around 860 s).
Advective transport by the mean velocity $\overline{U}=2.4$ \textmu m/s over the same
distance would happen on a timescale of $t_a=L/\overline{U}=83$ s, indicating that
transport to the posterior ventricle is solely driven by diffusion at the developmental
stage 2 dpf. The estimated timescale of diffusion $t_d=L^2/D_1$ over the 
same distance is $\approx$24500 s. Larger molecules
transported primarily by advection are thus more easily
compartmentalized within the ventricles
as a result of the compartmentalized CSF flow.

Removing motile cilia slowed down transport within
the middle ventricle and towards the anterior ventricle,
but the transport towards the posterior ventricle was unaffected.
Photoconversion experiments on cilia mutant
zebrafish larva (\emph{smh}) aligned with these observations.
This further indicates that
transport from the middle to the posterior ventricle is
purely driven by diffusion at this developmental stage.
The experimental results provide not only confidence
in the predictive ability of the computational model,
but also quantify the impact of cilia motility
on intraventricular transport.
Interestingly, Yoshida~\emph{et al.} observed a drastic reduction
in CSF mixing in specific regions of the brain
ventricles when simulating reduced cilia motility~\cite{Yoshida2022EffectVentricles}.
\lyng{Since impaired brain 
clearance of tracers has been demonstrated in idiopathic normal pressure hydrocephalus
patients~\cite{Eide2020MagneticHydrocephalus},
cilia-enhanced mixing in the brain ventricles might be
fundamental to physiological transport within the brain.}

>>>>

\begin{abstract}
Cerebrospinal fluid (CSF) is integral to brain function, as the fluid
provides mechanical support for the brain and the flow of CSF helps
distribute nutrients, neurotransmitters and metabolites throughout the
central nervous system. CSF flow is partly driven by beating of motile
cilia, hair-like structures that line the ependymal cell layer of the
walls of the brain ventricles. Despite the physiological importance of
CSF flow, the underlying mechanisms of cilia-mediated flow and
transport in the brain ventricles remain yet to be comprehensively
resolved. This study analyzes and evaluates the role of motile cilia
in brain ventricular CSF flow and transport. The geometry of the
zebrafish larval brain ventricles, for which detailed knowledge of
cilia properties and CSF motion velocities and frequencies exists, was
used to develop finite element methods for modelling flow and
transport. The computational model is validated by in vivo experiments
that monitor transport of a photoconvertible protein secreted in the
brain ventricles. Our results show that while cilia contribute to
advection of large particles, diffusion plays a significant role in
the transport of small solutes. We also demonstrate how cilia location
and the geometry of the ventricular system impact solute
distribution. Altogether, this work presents a computational framework
that can be applied to other ventricular systems, together with new
concepts of how molecules are transported within the brain and its
ventricles.

@@

\noindent\small\hspace{.975cm}\textbf{Keywords} cerebrospinal fluid, brain ventricles, cilia, transport, finite elements

@@

% I4: How do we address the problem and what do we find out?
In this study, we combined computational and experimental approaches to investigate
the role of motile cilia in brain ventricular CSF flow and transport.
We model the flow and transport by partial differential equations and solve
them numerically with finite element methods to compute flow patterns and
solute concentration dynamics. A geometry representing the zebrafish larval brain ventricles
was used, for which we have detailed knowledge of cilia properties and CSF motion velocities
and frequencies~\cite{Olstad2019CiliaryDevelopment}. 
Notably, this early developmental stage allows us to simplify our model
by omitting CSF secretion, which starts at later stages~\cite{Jeong2024TheZebrafish},
while still being well-conserved with mammals~\cite{DGama2025MotileBrain,
Olstad2019CiliaryDevelopment, Ringers2020Role, DGama2021Diversity, jurisch2020radial}. 
The computational model is validated against experimental data by monitoring in vivo
the dynamics of a photoconvertible fluorescent protein Dendra2
expressed within the brain ventricles.
Our results show that diffusion plays a major role in the transport of small solutes,
whereas cilia are paramount for the movement of larger particles.
We also found that the location of cilia, as well as changes to the ventricular geometry,
largely impact solute distribution.
In summary, our work presents a computational
framework that can be applied to other
ventricular systems in the future, together with new concepts
of how cilia distribute molecules and
particles within the brain ventricles.

@@

Assuming that viscous forces are dominating the ventricular CSF flow,
we model the flow by the Stokes equations.
The net forces of motile cilia acting on the CSF are modeled as a steady,
bidirectional tangential traction acting on parts of the ventricular wall
(\Crefpanels{fig:fig1}{d}), with free velocity slip at the walls.
Simulations show that this traction alone
leads to a partial compartmentalization of the ventricular system,
with large-scale vortex structures in the anterior, middle and
posterior ventricles (\Crefpanels{fig:fig1}{e}). The highest flow
speeds ($26.7$ \textmu m/s) occur in the vicinity of the dorsal cilia
in the middle ventricle. To model the cardiac pulsations, we impose a
pulsatile pressure difference between the anterior and posterior
boundaries (\Crefpanels{fig:fig1}{d, f}). The resulting flow
is pulsatile and directional (\Crefpanels{fig:fig1}{g}), and
in the absence of the cilia, the vortex structures vanish. The
velocity magnitude peaks at $4.8$ \textmu m/s. CSF initially flows
rostrocaudally before changing direction to caudorostral in the middle
of the cardiac cycle.

@@

Ciliary motion is integral to physiological CSF flow,
and motile-cilia deficiency has been
associated with defective brain development and various
pathologies in zebrafish, rodents and humans~\cite{Eichele2020Cilia-drivenVentricle,
Olstad2019CiliaryDevelopment, Faubel2016Cilia-basedVentricles,
Guirao2010CouplingCilia, Afzelius2004CiliaRelatedDiseases,
Sawamoto2006NewBrain, Yoshiba2014RolesSymmetry,
Hirokawa2006NodalAsymmetry, Youn2018PrimaryDiseases,
Ma2024CiliaDisease, Ibanez-Tallon2002LossHydrocephalus}.  As observed both experimentally and
computationally, absence of motile cilia alters CSF flow patterns.  We
next investigate the implications of these changes on brain ventricular
transport. We experimentally measured the movement of proteins by
monitoring photoconverted Dendra2 signal in control zebrafish and
motile-cilia mutant \emph{schmalhans} (\emph{smh}) zebrafish.  In ROIs
2 and 4, which quantify Dendra2 distribution downstream of the
photoconversion site, we observed a slower increase of fluorescence
intensity (\Crefpanels{fig:fig4}{a}): the mean times-to-threshold in
the \emph{smh} mutants were $45\%$ (ROI 2) and $25\%$ (ROI 4) higher
compared to controls (\Crefpanels{fig:fig4}{b}). We found no
significant differences between \emph{smh} mutants and controls in the
Dendra2 distribution in ROIs 3, 5, and 6.

In the computational model, we represent the \emph{smh} mutants by
considering transport of Dendra2 with CSF flow driven by cardiac
pressure pulsations only, representing the absence of motile
cilia. Comparing the in silico mutants with transport predictions in
the presence of both cardiac pulsations and motile cilia, we observe
noticeably delayed dynamics in ROIs 2, 4 and 5, but little change for
ROIs 3 and 6 (\Crefpanels{fig:fig4}{c}). Compared to experimental
data, the control and mutant scenarios differed more for the
simulations, with the times-to-threshold in ROI 2 and 4 being 81\% and
28\% higher in the absence of cilia-mediated flow. Moreover, the
predicted mean concentrations in ROI 5 deviate between the two
scenarios, which is not the case in the experiments. Altogether, our
findings identify that, for smaller proteins such as Dendra2,
diffusion plays an important role for solute movement to regions
upstream in the cilia-mediated flow fields or to distant ventricles,
while the cilia affect the downstream mean concentrations.
\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{graphics/figure4_compare_control_mutant.png}
    \caption{\textbf{a.} Relative change in fluorescence intensity $\Delta F(t)$
    in each region of interest (ROI) for control (black lines)
    and \emph{smh} mutant (red lines) zebrafish. 
    Bold lines are mean values of the cohorts and shaded regions cover one standard devation.
    \textbf{b.} Times when $\Delta F(t)$ first exceeds a threshold value,
    equal to 0.25 (ROIs 2--4) or 0.10 (ROIs 5, 6), for controls (gray dots) and
    \emph{smh} mutants (red dots).
    The thick black lines denote mean values. The $p$-values
    were calculated using a Mann-Whitney U test with 95\% confidence level,
    under the null hypothesis that the two distributions of control and mutant data are 
    from the same population.
    \textbf{c.} Mean concentration $\cbar(t)$ in each ROI,
    simulated using the baseline model (black lines) and the cardiac-induced/no-cilia model (red lines).
    The horizontal gray lines mark the threshold value $\hat{c}$, and the time
    $\hat{t}$ when the threshold is first exceeded is reported in the plot legends.
    Note that in panels \textbf{a} and \textbf{c} the vertical axes have
    different limits in each plot.}
    \label{fig:fig4}
\end{figure}

@@

We previously characterized three distinct ciliated cell lineages in
the zebrafish larval brain~\cite{Olstad2019CiliaryDevelopment,
  DGama2025MotileBrain}.  To identify the effects of these lineages on
CSF flow and transport, we investigate region-specific cilia paralysis
with the computational model.  Three cilia-covered regions were
paralyzed separately: the anterior ventricle, and the dorsal and
ventral parts of the middle ventricle (\Crefpanels{fig:fig5}{a}).  To
simulate regional cilia paralysis, we remove the tangential traction exerted by
the cilia in the respective region.

Removing cilia in the dorsal part of the middle ventricle has a strong
effect on the CSF velocities: the maximum velocity magnitude drops
from 27.9 \textmu m/s to 7.9 \textmu m/s. In vivo, these dorsal cilia
generate the strongest flow at this developmental stage of the
fish~(\Crefpanels{fig:fig1}{c})~\cite{Olstad2019CiliaryDevelopment}.
In contrast, only removing cilia in the anterior ventricle or the
ventral region of the middle ventricle has minimal impact, and results
in maximum speeds of 27.9 \textmu m/s and 27.4 \textmu m/s,
respectively.  We remark that out of the total tangential forces,
ventral cilia in the middle ventricle contribute by 34\%, while the
dorsal cilia exert 52\% of the total tangential
force~(\Crefpanels{fig:fig5}{b}).
Thus, the CSF velocities are affected not only by
the magnitude of the applied forces, but result
from a nontrivial interplay between force directionality
and location, and ventricular morphology.

Turning to transport of Dendra2, removal of the
dorsal cilia again has the most pronounced effect. All mean concentration
profiles (except ROI 1, where we impose the photoconversion curve)
change noticeably (\Crefpanels{fig:fig5}{c}). Compared to the baseline
model results, we observe less solute distribution in the middle and
anterior ventricles (ROIs 2--5), and more rapid transport to the
posterior ventricle (ROI 6). On the other hand, removal of anterior
cilia does not alter the mean concentrations
(\Crefpanels{fig:fig5}{c}). Finally, when removing ventral cilia in the
middle ventricle more solute spreads towards the anterior ventricle
(ROI 5), and less solute spreads towards the posterior ventricle (ROI 6).
\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{graphics/figure5_compare_cilia_modifications.png}
    \caption{\textbf{a.} The three separate cilia regions for which cilia paralysis is simulated
    by imposing a zero tangential traction boundary condition in the respective region.
    \textbf{b.} The total tangential force applied to the boundary with the tangential traction
    boundary condition in each of the three separate ciliated regions, as well as the total force
    (the sum of the three regions), calculated with \Cref{eq:tangential_force_calculation}. 
    \textbf{c.} Mean concentration $\cbar(t)$ for the six regions of interest (ROIs) in the three 
    paralysis scenarios considered. The region named in the legend is the region
    where we modify the cilia tangential traction boundary condition to a free-slip condition
    in the CSF flow model.}
    \label{fig:fig5}
\end{figure}

\subsection{Modification of ventricular geometry impacts solute distribution}
Variations in ventricular morphology impact CSF flow
patterns, and we previously identified large
interindividual variation in ventricular size and
CSF velocities between larva at given developmental stages and throughout
development~\cite{Olstad2019CiliaryDevelopment, DGama2025MotileBrain, Jeong2024TheZebrafish}.
To assess how these
variations influence solute distribution, we consider four alternative
ventricular geometries: (i) constriction of the anterior-middle
interventricular duct (66 \% reduction in cross-sectional area);
(ii) constriction of the middle-posterior duct (33 \% reduction);
(iii) middle ventricle shrunk in the lateral direction (43 \% reduction);
and (iv) all three combined (\Crefpanels{fig:fig6}{a, b}).
In these four modified geometries, the total volume is reduced by
3.9\%, 4.1\%, 1.4\% and 9.4\%, respectively, compared to the original
geometry. As a result of the geometry modifications,
the forces applied with the 
cilia tangential traction differ across the scenarios
(\Crefpanels{fig:fig6}{c}).

Constricting the anterior-middle duct
reduces solute transport to the anterior ventricle (ROI 5), and the
mean concentrations in ROIs 2--4 and 6 slightly increase, compared to
the original geometry (\Crefpanels{fig:fig6}{d}).
Constricting the middle-posterior duct
mainly delays the transport into the posterior ventricle (ROI 6),
while slightly accelerating transport towards the anterior
ventricle (ROI 5). 
Thus for both scenarios, solute transport is restricted through
the respective constricted interventricular duct.
For the geometry with a shrunk middle
ventricle, we observe faster transport
to the anterior ventricle (ROI 5)
and slower transport to the lower middle 
and posterior ventricles (ROI 3, ROI 6). 
Interestingly, this reduction in the transport to the
posterior ventricle (ROI 6) happens in spite of a 6.6\% increase in
the ventral cilia forces, which are the cilia that propel
CSF towards the posterior ventricle.

When combining all three modifications,
we observe higher retention of solute in the middle ventricle (ROIs 2--4),
while the mean concentrations are reduced in both the anterior
and the posterior ventricles (ROIs 5 and 6).
To summarize, these results show that the ventricular geometry and the
cross-sectional area of the ventricular ducts clearly impact solute
transport, with notable differences in transport resulting from modest
geometry modifications. These findings may explain the larger
differences in the cilia-deficient computational model results
compared to the differences in experimental data in \Cref{fig:fig4},
as the computational geometry is not an exact representation of the
individual zebrafish ventricles.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{graphics/figure6_compare_modified_geometries.png}
    \caption{\textbf{a.} The original computational mesh of the ventricular geometry.
    \textbf{b.} The four modified versions of the computational geometry. Compared
    to the original mesh, the volumes of the deformed meshes are reduced by 3.9\%, 4.1\%, 1.4\% 
    and 9.4\%.
    \textbf{c.} Mean concentration $\cbar(t)$ simulated with the original and modified geometries
    in the six regions of interest (ROIs).
    \textbf{d.} A comparison of the tangential cilia force in the three cilia regions
    and the total force applied, calculated with \Cref{eq:tangential_force_calculation}.
    In each region, the force $F_{\mathrm{original}}$ 
    applied on the original mesh is the baseline of 100\%,
    to which the forces $F_i$ of the other mesh versions are compared, where $i$ denotes
    the modification scenario.
    The bars for the four alternative scenarios display $F_i/F_{\mathrm{original}}\times 100\%$.}
    \label{fig:fig6}
\end{figure}

@@

Using the simulated CSF flow, we modeled solute transport by an advection-diffusion equation.
The transport results compared well with data from in vivo experiments in zebrafish.
Then, by using a computational model as well as zebrafish mutants with paralyzed cilia,
we identified that even though advection by cilia is dominant,
diffusion plays a significant role in protein transport.
Finally, we investigated how local cilia paralysis and ventricular geometries
affect solute distribution. Altogether, we present not only new findings on the roles of
advection and diffusion in brain ventricular transport, but also provide a
computational framework to study CSF flow and transport in ventricular geometries.

\subsection{The roles of advection and diffusion in brain ventricular solute transport}
To investigate brain ventricular solute transport, we quantified transport
based on changes in fluorescence intensity $\Delta F$ (experiments) and
mean concentration $\cbar$ (simulations) over time. Our results convincingly show that
the solute transport depends on a balance of advection and diffusion.
While advection is the strongest transport mechanism on a global scale for
the three solutes we studied (global Péclet numbers greater than unity),
diffusion is balancing out advection for shorter distances and in regions without motile cilia activity.
For instance, for the larger extracellular vesicles, no solute reached the (distant)
posterior ventricle (a distance of $L = 200$ \textmu m) within the simulation time (around 860 s).
Advective transport by the mean velocity $\overline{U}=2.4$ \textmu m/s over the same
distance would happen on a timescale of $t_a=L/\overline{U}=83$ s, indicating that
transport to the posterior ventricle is solely driven by diffusion at this developmental
stage. The estimated timescale of diffusion $t_d=L^2/D_1$ over the 
same distance is $\approx$24500 s, compared to 696 s and 348 s for the 
smaller proteins (STM-GFP and Dendra2).
This explains why $D_1$ is the only
case where no solute spreads to the posterior ventricle,
while STM-GFP and Dendra2 are transported there by diffusion.
Larger molecules transported primarily by advection
are thus more easily
compartmentalized within the ventricles
as a result of the compartmentalized CSF flow.

\subsection{Effects of cilia motility on transport}
Removing motile cilia slowed down transport within
the middle ventricle and towards the anterior ventricle,
but the transport to the posterior ventricle was unaffected.
Photoconversion experiments with cilia mutant
zebrafish larva (\emph{smh}) aligned with these observations.
This further indicates that
transport from the middle to the posterior ventricle is
purely diffusive at this developmental stage,
while transport within the middle ventricle
depends on cilia motility.
Interestingly, Yoshida~\emph{et al.} observed a drastic reduction
in CSF mixing in specific regions of the brain
ventricles when simulating reduced cilia motility~\cite{Yoshida2022EffectVentricles}.
Our experimental results provide not only confidence
in the predictive ability of our computational model,
but also quantify the impact of cilia motility
on brain ventricular transport.

\subsection{Model parameters and sources of discrepancies}
Even though we identified similar dynamics in simulations and
in vivo experiments, the transport dynamics of the simulations were generally
more rapid and resulted in higher total amounts of solute.
Differences between experiments and simulations may relate to
ventricular size, as we showed that modifying the geometry of the
ventricular geometry impacts solute distribution,
and the ventricular size varies substantially 
in the control population at 2 dpf.
Additionally, differences may result
from the definition of the photoconversion site,
since the volume of this region determines the total amount of solute
and thereby the rate of transport. Another source
of discrepancy may relate to the permeability of the ventricular wall,
which we considered as impermeable in our simulations.
The neuroepithelium lining the walls of brain ventricles in 24 hour old zebrafish
is permeable to dyes smaller than 70 kDa~\cite{Chang2012AnNeuroepithelium}.
If this is also the case at 2 dpf, Dendra2 
(which are 26 kDa in size~\cite{Gurskaya2006EngineeringLight})
would leak out of the brain ventricles over the relevant time spans
and render the dynamics slower.
It may also be the case that our model predicts faster
transport because the Dendra2 diffusion
coefficient we used (based on an experimental value
\cite{GuraSadovsky2017MeasurementExpansion}) is too high.
This could be because embryonic CSF has higher viscosity
than the solution considered in the measurements, 
or because the presence of proteins, nutrients and other solutes
obstructs diffusive transport
\cite{Bloomfield1998EffectsFluid, Einstein1905UberTeilchen, Cussler2009Diffusion:Systems}.
Finally, an important difference
is the three-dimensional nature of the computational model,
while the experimental images only show 2D slices.
